---
title: "BOHA Veg Monitoring"
output:
  html_document:
    toc: true
    toc_float: true
---

# Load in packages and monitoring data

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load packages
library(tidyverse)
library(magrittr)
library(lubridate)
library(readxl)
library(knitr)
library(ggpubr)

#### import data sets and save as lists to working directory----

#Thompson Island

# Thompson 2019
path<-"./data/Thompson2019.xlsx" 

Thomp19<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)
  
# Thompson 2018
path<-"./data/Thompson2018.xlsx" 

Thomp18<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

# Thompson 2017
path<-"./data/Thompson2017.xlsx"

Thomp17<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

# # Thompson 2016
path<-"./data/Thompson2016.xlsx"

Thomp16<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

# Thompson 2015
path<-"./data/Thompson2015.xlsx"

Thomp15<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)


# Grape

# Grape 2019
path<-"./data/Grape2019.xlsx" 

Grape19<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

# Grape 2018
path<-"./data/Grape2018.xlsx" 

Grape18<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

# Grape 2017
path<-"./data/Grape2017.xlsx"

Grape17<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

# Grape 2016
path<-"./data/Grape2016.xlsx" 

Grape16<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

# Grape 2015
path<-"./data/Grape2015.xlsx" 

Grape15<-path %>%
  excel_sheets() %>%
  purrr::set_names() %>%
  map(read_excel, path = path)

str(Thomp15)
```

# Herb Layer
## Data Checks: combine  data, grouping variables, missing values, and sampling events

```{r herbchecks, echo=TRUE}

# Clean up column values (Island and Zone) and Derive new variables from the data set: year, month, Pct_Cover

### Extract annual data from layer sampling and  combine annual data per subprotocol (e.e.g, herb layer) ----


# Extract layer data per year/island and combine into one DF. Also currently removes the notes fields for easier binding (1 or more years without NOtes w/ ID field)

herbs<-bind_rows(Thomp15$`HERB LAYER` %>%  dplyr::select(-starts_with("Note"), -First_point_cm), Thomp16$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")),Thomp17$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")), Thomp18$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")),Thomp19$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")), Grape15$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")),Grape16$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")), Grape17$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")),Grape18$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")),Grape19$`HERB LAYER` %>%  dplyr::select(-starts_with("Note")))


herbs <-herbs %>% mutate(year= year(Date), month= month(Date)) %>% # extract year and month for later filtering/modelling
  mutate(Pct_Cover = case_when( 
    Cover == 0  ~ 0,
    Cover == 1  ~ 0.25,
    Cover == 2  ~ 0.5,
    Cover == 3  ~ 1.5,
    Cover == 4  ~ 3.5,
    Cover == 5  ~ 7.5,
    Cover == 6  ~ 17.5,
    Cover == 7  ~ 37.5,
    Cover == 8  ~ 62.5,
    Cover == 9  ~ 85,
    Cover == 10  ~ 97.5,
    TRUE ~ NA_real_)) %>%   # create median cover class for each estimate of Cover: 1= <0.5%;  2 = 0-1%;  3 = 1-2%;  4 = 2-5%;  5 = 5-10%;  6 = 10-25%;  7 = 25-50%;  8 = 50-75%;  9 = 75-95%;  10 = 95-100%
  mutate(Island = case_when(Island == "THOMPSON" ~ "Thompson",
                            Island == "Thompson" ~ "Thompson",
                            Island == "GRAPE" ~ "Grape",
                            Island == "Grape" ~ "Grape")) %>% # rename Islands for consistency
  mutate(Zone = case_when(Zone == "U" ~ "Upland",
                          Zone == "W" ~"Wetland",
                          Zone == "Wet" ~"Wetland",
                          Zone == "UP" ~"Upland",
                          Zone == "WET" ~ "Wetland",
                          Zone == "C" ~ "Control",
                          Zone == "Upland" ~ "Upland",
                          Zone == "Control" ~ "Control",
                          Zone == "Wetland" ~ "Wetland")) %>%  # rename zones for consistency
  filter(!Quadrat %in% 0 ) %>% # remove quadrats labeled 0 which denote an entry of a transect note and not obs
  filter(!Species %in% "****" ) %>% # remove records created for note taking
  filter(!Species %in% "-" )



# check the levels or values of a few fields

unique(herbs$Island)# unique values of island
unique(herbs$Zone)


kable(herbs[1:51,])
```

### Check for missing values
```{r}

# Are there any missing values that need to be addressed:

# in the year column
kable(herbs[is.na(herbs$year),], caption = "Records with missing values in the year column")

# in the Island column
kable(herbs[is.na(herbs$Island),], caption = "Records with missing values in the Island column")


# in the Quad column
kable(herbs[is.na(herbs$Quadrat),], caption = "Records with missing values in the Quadrat column")

# Species with NA
## there are apparently some species/cover classes that don't have cover values assigned to them.

kable(herbs[is.na(herbs$Cover),], caption = "Records without cover class data")

#### SHOULD WE DROP THESE SPECIES?

#These species will be removed from an analysis of cover but included in the an analysis of quad freq. 

write_csv(filter(herbs, is.na(Cover)), "./data/checks/herbspeciesNoCover.csv")

# There are entries in the data without species codes

kable(herbs[is.na(herbs$Species),], caption ="Records without species codes but with cover data")

```

### Check the number of sampling events per island, zone, transect and year

```{r herb_obs}

#Count number of Transects sampled per Zone per year

herb.trans.year<-herbs %>% dplyr::select(Island, year, Transect,  Zone) %>%
  group_by(Island, year, Zone) %>% 
  distinct() %>% 
  tally(name= "Transects_Sampled") %>% pivot_wider( names_from= year, values_from = Transects_Sampled)

kable(herb.trans.year, caption = "Number of Transects sampled per Zone per year. NA indicates missing data.")


# Are there sampling events at each transect in each year?

herb.obs.trans.Zone<-herbs %>% dplyr::select(Island, Transect, Zone, year) %>%
  group_by(Island, Transect,  Zone, year) %>% 
  distinct() %>% arrange(desc(-year)) %>% pivot_wider( names_from= year, values_from= year) %>% arrange(Island, Transect)

kable(herb.obs.trans.Zone, caption = "Transects sampled each year. NA indicates missing data.")

write_csv(herb.obs.trans.Zone,"./data/checks/herb.transects.year.csv")

```

### Check the number of quadrats sampled events per island, zone, transect and year

```{r}
#Count number of quads  sampled per transect, Zone per year

herb.quads.year<-herbs %>% dplyr::select(Island, year, Zone, Transect,Quadrat) %>%
  group_by(Island, year, Zone, Transect) %>% 
  distinct() %>% 
  tally(name= "Quads_Sampled") %>% pivot_wider(names_from = year, values_from = Quads_Sampled) %>% arrange(Island, Transect)


# there should be 20 quads per transect but there are some cases with 19, 21, or 22 quads sampled? I will use the ctual number of quads sampled to calc means becuase there are some transects with 21 quads with data entered. Some of these numbers will be changed by removing Species == ****.



kable(herb.quads.year, caption= "Number of quadrats sampled per transect in each year. NA indicates missing data.")

write_csv(herb.quads.year,"./data/checks/herb.quads.year.csv")
```


### Check duplicate transect sampling in Grape 2017

In 2017 transects GC 58 and 59 were sampled twice using different quads

```{r, GC58_59}

trans58<-filter(herbs, Transect %in% c("GC58.1", "GC58.2")) %>% group_by(Transect, Species) %>% 
  summarise(MeanCover = mean(Pct_Cover, na.rm= TRUE)) %>% 
  pivot_wider(names_from = Transect, values_from=MeanCover) %>% 
  mutate(mean = rowMeans(.[,2:3]))
  
  

kable(trans58, caption ="Mean cover by Species for duplicate transect GC58")


trans59<-filter(herbs, Transect %in% c("GC59.1", "GC59.2")) %>% group_by(Transect, Species) %>% 
  summarise(MeanCover = mean(Pct_Cover, na.rm= TRUE)) %>% 
  pivot_wider(names_from = Transect, values_from=MeanCover)%>% 
  mutate(mean = rowMeans(.[,2:3]))


kable(trans59, caption ="Mean cover by Species for duplicate transect GC59")

# rename  transects sampled > 1 per year in herbs object to combine and avg transect data

herbs<-herbs %>% 
 mutate(Transect = case_when(Transect == "GC58.1" ~ "GC58",# in 20-17 there are two transects sample twice, renaming to main transect name to avg cover
                              Transect == "GC58.2" ~ "GC58",
                              Transect == "GC59.1" ~ "GC59",
                              Transect == "GC59.2" ~ "GC59",
                              TRUE ~ as.character(Transect)))
 
 
```




### Generate list of observed species codes

```{r, herbspecies}
herb.species<-herbs %>%  distinct(Species) %>% as_tibble() # generate list

paste(tally(distinct(herb.species,Species)), "species observed!")
       
       
# import species list provided by Rachel Vincent: plant_lists_GRAP_THOM_FieldGiude.xlsx

species_tlu <- read_excel("data/plant_lists_GRAP_THOM_FieldGiude.xlsx")

paste(tally(distinct(species_tlu,`BOHA Code`)), "species in field guide")

#Return the list of observed species from the field guide

Obs.Herbs<-semi_join(species_tlu,herb.species, by = c("BOHA Code"= "Species")) 

# Show the list of observed species NOT in the field guide

paste(tally(distinct(anti_join(species_tlu,herb.species, by = c("BOHA Code"= "Species")))), "species not in field guide")

kable(anti_join(species_tlu,herb.species, by = c("BOHA Code"= "Species")), caption ="List of observed species not in the field guide")
```


## Population estimation

```{r species_tlu, echo= F}
# import tables with species of interest (from RSV 7/31/21)

TopSpecies <- read_excel("./data/Plants_TopSpecies.xlsx") 

# Check species not matching with this Top Species list

sppCheck<- herbs %>%  left_join(.,TopSpecies, by="Species") %>% 
  filter(is.na(Origin))  %>%  pull(Species) %>% unique() %>% as_tibble()
  
write_csv(sppCheck,"./data/checks/SpeciesCodesNotMatched.csv")

kable(TopSpecies[,1:4], caption ="List of Species of interest")
```

### <b> Quadrat frequency </b>

Quadrat frequency:  the percentage of quadrats per transect containing at least one species; tracks the local establishment and expansion of a species within a transect.

The following code is used to calculate the mean proportion of quadrat occupied per transect for each island, zone and year. Species occupancy is calculated first at the transect level (# quads present /total quads sampled) and then averaged across transects for a zone-scale average.

#### Data checks: Species occurence
```{r, herbQuadCheck}

##### QUAD FREQUENCY

# count the number of quads sampled per transect in each year

herb.quads.year<-herbs %>% dplyr::select(Island, year, Zone, Transect,Quadrat) %>%
  group_by(Island, year, Zone, Transect) %>% 
  distinct() %>% 
  tally(name= "Quads_Sampled")

#  DETERMINE THE FREQ EACH SPECIES was DETECTED EACH YEAR

herb.species.TransFreq <-herbs %>%  group_by(Island, Zone, year, Transect, Species) %>% tally(name = "Quad_Present") %>% # count the number of species occurences
    left_join(herb.quads.year, by = c("Island", "Zone", "year", "Transect")) %>%   # ADD ON NUMBER OF QUADS SAMPLED PER YEAR (FROM ABOVE)
    mutate(Trans_Freq = Quad_Present/Quads_Sampled) # calc the proportion of quads within a transect containing that species

#Check potential double entry of Species Codes per quadrat:

herb.check.dups<-herbs %>%  group_by(Island, Zone, year, Transect, Quadrat, Species) %>% tally(name ="Detected") %>% filter(Detected >1)

kable(herb.check.dups, caption =" List of records where species was entered >1 in same quadrat. Will need to search data to determine which quadrats ")


write_csv(herb.check.dups, "./data/checks/herb.check.dups.csv")

# Across all quads

herb.MeanSiteQuadFreq <-herbs %>%  group_by(Island, Zone, year, Transect, Species) %>% tally(name= "Quad_Present") %>% 
  left_join(herb.quads.year, by = c("Island", "Zone", "year", "Transect")) %>%   # ADD ON NUMBER OF QUADS SAMPLED PER YEAR (FROM ABOVE)
  mutate(Trans_Freq = Quad_Present/Quads_Sampled) %>% 
  group_by(Species) %>% 
  summarise(TransFreq = round(mean(Trans_Freq, na.rm= TRUE),3), Quads_Present= n()) %>% 
  arrange(desc(TransFreq)) %>% filter(Quads_Present > 20)

kable(herb.MeanSiteQuadFreq[1:20,], caption =" 20 most commonly detected species (detected in a minimum of 20 quads across study")

write_csv(herb.MeanSiteQuadFreq, "./data/checks/herb.MeanSiteQuadFreq.csv")

# Trans Freq by By Zone

herb.MeanTransFreqZone <-herbs %>%  group_by(Island, Zone, year, Transect, Species) %>% tally(name= "Quad_Present") %>% # count the number of quads the species was detected in 
  left_join(herb.quads.year, by = c("Island", "Zone", "year", "Transect")) %>%   # ADD ON NUMBER OF QUADS SAMPLED PER TRANSECT PER YEAR (FROM ABOVE)
  mutate(Trans_Freq = Quad_Present/Quads_Sampled) %>% 
  group_by(Zone, Species) %>% 
  summarise(MeanTransFreq = round(mean(Quad_Present, na.rm= TRUE),2), Quads_Present= n()) %>%  # cALCULATE SUMMARY STATS OF MEAN SPECIES QUAD FREQ
  arrange(desc(MeanTransFreq)) %>% filter(Quads_Present > 20) %>% 
  pivot_wider(id_cols= Species,names_from = Zone, values_from = MeanTransFreq)

kable(herb.MeanTransFreqZone[1:35,], caption =" 35 most commonly detected species (Avg transect freq in a minimum of 20 quads across study")


```

#### Calculate Mean Quad frequency

Calculate the proportion of quads each species has been detected per Zone and island
```{r herbfreq}

# count the number of quads sampled per transect in each year

herb.quads.year<-herbs %>% dplyr::select(Island, year, Zone, Transect,Quadrat) %>%
  group_by(Island, year, Zone, Transect) %>% 
  distinct() %>% 
  tally(name= "Quads_Sampled")

# reduce data to species of interest 

herbsSpecies<-herbs %>% left_join(TopSpecies[,1:4],.,by = "Species")  #bind sci names and extract data of interest

# Complete data frame of the transect sampling by island/zone and calc Quad freq per Zone
# i.e. add 0 for species not observed in certain quads

herb.quadfreq<-herbsSpecies %>% group_by(Island, Zone, year, Transect, `USDA Plants Code`) %>% tally(name = "Quads_Present") %>% ungroup() %>% 
    complete(`USDA Plants Code`, nesting(Island, Zone, Transect, year), fill = list(Quads_Present = 0)) %>%  # create 0 records for each species when it was not observed for each transect island combo
  drop_na() %>% # for some reason the above code is creating NS records, this drops them and doesn't seem to effect design
  left_join(herb.quads.year, by = c("Island", "Zone", "Transect","year"))# add the specific no. of quads sampled each year per transect

# check to ensure consistent quad sampling by species

transcheck<-herb.quadfreq %>% select(`USDA Plants Code`, Island, year, Zone, Transect) %>% 
  group_by(`USDA Plants Code`, Island,  Zone, year ) %>% 
  distinct() %>% 
  tally(name= "Trans_Sampled") %>% pivot_wider(names_from = year, values_from = Trans_Sampled) %>% arrange( Island, Zone, `USDA Plants Code`)

kable(transcheck, caption =" No. of transects sampled per species")

quadcheck<-herb.quadfreq %>% select(`USDA Plants Code`, Island, year, Zone, Quads_Sampled) %>% 
  group_by(`USDA Plants Code`, Island,  Zone, year ) %>% 
  summarise(Quads_Sampled = sum(Quads_Sampled)) %>% 
  pivot_wider(names_from = year, values_from = Quads_Sampled) %>% arrange( Island, Zone, `USDA Plants Code`)


kable(quadcheck, caption =" No. of quads sampled per species")


# calculate avg proportion of quads per transects in each year by species

 herbs.AvgQuadFreq<- herb.quadfreq %>% 
 mutate(PropQuadsPerTrans = Quads_Present/Quads_Sampled) %>%   # calc the proportion of quads within a transect containing each species
group_by( `USDA Plants Code`,Island, Zone, year) %>%  
  summarise(MeanPropPerZone = mean(PropQuadsPerTrans, na.rm = TRUE), SDPropZone = sd(PropQuadsPerTrans, na.rm = TRUE), `Transects Sampled` = n(), SEPropZone= sd(PropQuadsPerTrans, na.rm = TRUE)/sqrt(n()) ) %>% 
left_join(distinct(dplyr::select(TopSpecies, `USDA Plants Code`, `Scientific name`, `Common Name`, Origin)),by = "USDA Plants Code")

 write_csv(herbs.AvgQuadFreq,"./data/HerbFreqPerZone.csv")
 
```

### Plot Quad frequency

```{r plot_herb_Freq, comment= FALSE, echo=FALSE, message=FALSE}


species<-unique(TopSpecies$`USDA Plants Code`)

freq_plot = function(df, species) {
  
  plotdata<-dplyr::filter(df, `USDA Plants Code` %in% species) %>% droplevels()

  ggplot(plotdata , aes(x= year, y = MeanPropPerZone, colour= Zone)) + geom_point() +geom_line()+
  geom_errorbar(aes(ymin= MeanPropPerZone-SEPropZone, ymax = MeanPropPerZone+SEPropZone, colour= Zone))+
  #scale_color_viridis_d()
  scale_x_continuous(breaks= seq(2015,2019,2))+
  facet_wrap(~Island)+
  theme_classic()+
  labs(title = paste(plotdata$`Common Name`[plotdata$`USDA Plants Code` %in% species], " - ", plotdata$Origin[plotdata$`USDA Plants Code` %in% species]), y= "Mean proportion of quadrats +- SE", x="Year")
  
}

Freq_plots = map(.x= species, ~freq_plot(herbs.AvgQuadFreq, .x))

freq_plots<-ggpubr::ggarrange(plotlist =Freq_plots, nrow=1, ncol =1, align = "hv", common.legend = TRUE, label.x = "", label.y = )

print(freq_plots)

ggpubr::ggexport(plotlist = freq_plots, filename = "BOHA_FreqPlots.pdf")


```


### Table of Mean Freq values

```{r herbFreqtable}

# The data table of avg freq

kable(herbs.AvgQuadFreq, caption = "Plotted herb quad freq. data")

```


### <b> Cover (abundance) </b>

#### Create dataframe for calculating cover and check sampling per species

The following code is used to calculate the mean cover per species for each island, zone and year. Cover is averaged first at the transect level and then across transects for a zone-scale average.

``` {r, herbCovercheck}
#bind sci names and extract data of interest
herbsSpecies<-herbs %>% left_join(TopSpecies[,1:4],.,by = "Species")  

# add in 0 records for each species not observed within transect-quad island combos

herb.Cover<-herbsSpecies %>% 
  select(`USDA Plants Code`, Island, Zone, year, Transect, Quadrat, Pct_Cover) %>%
   complete(`USDA Plants Code`, nesting(Island, Zone, Transect, Quadrat, year), fill = list(Pct_Cover = 0)) %>%  
  drop_na() 


# check to ensure consistent quad sampling by species

transcheck<-herb.Cover %>% select(`USDA Plants Code`, Island, year, Zone, Transect) %>%
  group_by(`USDA Plants Code`, Island,  Zone, year ) %>%
  distinct() %>%
  tally(name= "Trans_Sampled") %>% pivot_wider(names_from = year, values_from = Trans_Sampled) %>% arrange( Island, Zone, `USDA Plants Code`)

kable(transcheck, caption =" No. of transects sampled per species")

quadcheck<-herb.Cover%>% select(`USDA Plants Code`, Island, year, Zone, Transect) %>%
  group_by(`USDA Plants Code`, Island,  Zone, year, Transect ) %>%
  tally(name = "Quads_Sampled") %>%
  pivot_wider(names_from = year, values_from = Quads_Sampled) %>% arrange( Island, Zone, `USDA Plants Code`)


kable(quadcheck, caption =" No. of quads sampled per species. All quads should be 20 sampled with a few transects with only 19 sampled quads.")

```

#### Calculate avg cover 

```{r herbCoverCalc}

# calculate the mean cover of each species overtime at the Zone scale

 herbs.AvgCover<- herb.Cover %>% # use completed data set from above
   group_by( `USDA Plants Code`, Island, Zone, year, Transect) %>% # summarize data by the denoted columns
   summarise(MeanCover = mean(Pct_Cover, na.rm=TRUE), SDCover = sd(Pct_Cover, na.rm=TRUE), Quads_Sampled=n()) %>%   # calc the mean cover per transect for each species
group_by( `USDA Plants Code`,Island, Zone, year) %>% # now summarize data by Zone for each year
  summarise(MeanCoverPerZone = mean(MeanCover, na.rm = TRUE), SDCoverZone = sd(MeanCover, na.rm = TRUE), `Transects Sampled` = n(), SECoverPerZone = sd(MeanCover, na.rm = TRUE)/sqrt(n()) ) %>%
left_join(distinct(dplyr::select(TopSpecies, `USDA Plants Code`, `Scientific name`, `Common Name`, Origin)),by = "USDA Plants Code")

   write_csv(herbs.AvgCover,"./data/HerbCoverPerZone.csv") 
   
```

### Plot Mean Cover

```{r plot_herb_Cover, comment = FALSE, echo= FALSE, message=FALSE}

species<-unique(TopSpecies$`USDA Plants Code`)

cover_plot = function(df, species) {
  
  plotdata<-dplyr::filter(df, `USDA Plants Code` %in% species) %>% droplevels()

  ggplot(plotdata, aes(x= year, y = MeanCoverPerZone, colour= Zone)) + geom_point() +geom_line()+
  geom_errorbar(aes(ymin= MeanCoverPerZone-SECoverPerZone, ymax = MeanCoverPerZone+SECoverPerZone, colour= Zone))+
  scale_x_continuous(breaks= seq(2015,2019,2))+
  facet_wrap(~Island)+
  theme_classic()+
  labs(title = paste(plotdata$`Common Name`[plotdata$`USDA Plants Code` %in% species], " - ", plotdata$Origin[plotdata$`USDA Plants Code` %in% species]), y= "Mean cover (%) +- SE", x="Year")
}

Cover_plots = map(.x= species, ~cover_plot(herbs.AvgCover, .x))

coverplots<-ggpubr::ggarrange(plotlist =Cover_plots, nrow=1, ncol =1, align = "hv", common.legend = TRUE)


print(coverplots)

ggpubr::ggexport(plotlist = coverplots, filename = "BOHA_CoverPlots.pdf")

```

### Table of Mean Cover values

```{r herbCovertable}

# The data table of avg cover

kable(herbs.AvgCover, caption = "Plotted herb cover data")

```

